Probabilistic Catalogs for Crowded Stellar Fields

### 2012-10-09 paper/ms.tex
Hogg comments

However, standard methods for constructing catalogs can have difficulty in some
challenging situations. For example, when multiple sources overlap partially
or completely, it can be difficult to determine how many sources are present,
and how much flux belongs to each source. In principle, uncertainty should be
taken into account. Instead of simply {\it estimating} the position and flux of
each object in the image, we should fully describe the fact that sometimes we
are not certain of the position and flux of each source.

Essentially, the creation of a catalog is an attempt to answer the question
``given the image we have obtained, what objects are present in the field and
what are their properties?''. This motivates a probabilistic (Bayesian)
approach to making catalogs. The output of such an approach would not be a
single answer to this question (i.e. a single catalog), but rather a
probability distribution over the space of possible catalogs.
Constructing such a probability distribution is
challenging for a number of reasons. Firstly, the number $N$ of objects in the
image (and that should be in the catalog) is itself unknown. Secondly, if $N$ is large, then the parameter space
of positions and properties (flux, size, etc) of the objects is also large.
This can cause Markov Chain Monte Carlo (MCMC) algorithms difficulties -- they
may take a long time to converge to the target posterior distribution over
catalogs.

Bayesian object detection has been implemented by \citet{2011MNRAS.415.3462F}
under the assumption of a known number of objects $N$.
This assumption is required for the
{\tt MultiNest} sampler \citep{multinest} to be applicable.
Using the results from the known $N$
run, it is possible (under certain circumstances) to reconstruct what the
results would have been if a variable-$N$ simulation had been used. However,
this will not work well in situations where there is significant confusion
(i.e. two or more sources overlap). What is really required is a variable
dimension model, where $N$ is an unknown quantity to be inferred from the data.
The computational implementation of these models will require tools such as
reversible jump Markov Chain Monte Carlo \citep{rjmcmc}.

\section{Bayesian Inference}
To quantitatively model uncertainties, Bayesian Inference is the appropriate
framework \citep{cox, jaynes, caticha}. Suppose there exist unknown parameters
(denoted
collectively by $\theta$) and we expect to obtain some data $x$. Our prior
state of knowledge about the parameters is modelled by a prior
probability distribution:
\begin{equation}
p(\theta)
\end{equation}.
Note that this is a very concise notation \citep{hogg}.
We also model how the parameters give rise to the data, via a generative model.
This is also known as a {\it sampling distribution}.
\begin{equation}
p(x|\theta)
\end{equation}
Despite the singular, the sampling distribution is actually a family of
probability distributions over the data space, one for each possible value
of $\theta$. Note that
the sampling distribution is also an assertion of a prior state of knowledge:
It models prior information about the fact that the data $x$ is connected to
the parameters $\theta$ in some way. Without this prior knowledge, learning is
impossible: there has to be some relationship between the parameters and the
data, otherwise new data tells you about nothing but itself.

When specific data $x^*$ has been obtained, our state of knowledge about $\theta$
gets updated from the prior distribution to the posterior distribution,
via Bayes' rule:
\begin{eqnarray}
p(\theta|x=x^*) &\propto& p(\theta)p(x|\theta)|_{x=x^*} \\
&=& p(\theta)\mathcal{L}(\theta; x)
\end{eqnarray}
The term $p(\theta)p(x|\theta)|_{x=x^*} = \mathcal{L}(\theta; x)$
is the {\it likelihood function}, which is the 
probability of obtaining the actual data set $x^*$ as a function of the
parameters\footnote{In the case that the sampling distribution is a probability
density, the likelihood is the probability density evaluated at the observed
data. This usually causes no problems, although one should be aware of the
Borel-Kolmogorov paradox \citep{jaynes}.}.
As suggested by the above notation, the likelihood function is obtained from the
sampling distribution with the actual data substituted in.

\section{The Specific Model for Stellar Fields}

>>>>

However, standard methods for constructing catalogs can have difficulty in some
challenging situations. For example, when multiple sources overlap partially
or completely, it can be difficult to determine how many sources are present,
and how much flux belongs to each source. In principle, uncertainty should be
taken into account. Instead of simply {\it estimating} the position and flux of
each object in the image, we should fully describe the fact that sometimes we
are not certain of the position and flux of each source. This uncertainty
should then be propagated into any inferences about the stellar population.

Essentially, the creation of a catalog is an attempt to answer the question
``given the image we have obtained, what objects are present in the field and
what are their properties?''. This motivates a probabilistic (Bayesian)
approach to making catalogs. The output of such an approach would not be a
single answer to this question (i.e. a single ``point estimate'' catalog), but rather a posterior probability distribution over the space of possible catalogs.
Constructing such a probability distribution is
challenging for a number of reasons. Firstly, the number $N$ of objects in the
image (and that should be listed in the catalog) is itself unknown. Secondly, if $N$ is large, then the parameter space
of positions and properties (flux, size, etc) of the objects is also large.
This can cause Markov Chain Monte Carlo (MCMC) algorithms difficulties -- they
may take a long time to converge to the target posterior distribution over
the space of catalogs.

Bayesian object detection (as this problem is sometimes called) has been implemented by \citet{2011MNRAS.415.3462F}
under the assumption of a known number of objects $N$.
This assumption is required for the
{\tt MultiNest} sampler \citep{multinest} to be applicable.
Using the results from the known $N$
run, it is possible (under certain circumstances) to reconstruct what the
results would have been if a variable-$N$ simulation had been used. However,
this will not work well in situations where there is significant confusion
(i.e. two or more sources overlap). What is really required is a variable
dimension model, where $N$ is an unknown quantity to be inferred from the data.
The computational implementation of these models will require tools such as
reversible jump Markov Chain Monte Carlo \citep{rjmcmc}.

\section{Bayesian Inference}
To quantitatively model uncertainties and transform noise in observed data
into uncertainties in parameters of interest, Bayesian Inference is the appropriate
framework \citep{cox, jaynes, caticha}. Suppose there exist unknown parameters
(denoted
collectively by $\theta$) and we expect to obtain some data $x$. Our prior
state of knowledge about the parameters is modelled by a prior
probability distribution:
\begin{equation}
p(\theta).
\end{equation}
Note that this is a very concise notation \citep{hogg} and should be read
as ``the probability distribution for $\theta$''.
We also model how the parameters give rise to the data, via a generative model.
This is also known as a {\it sampling distribution}.
\begin{equation}
p(x|\theta)
\end{equation}
Despite the singular, the sampling distribution is actually a family of
probability distributions over the space of possible data, one probability
distribution for each possible value
of $\theta$. Note that the choice of
the sampling distribution is also an assumption about prior knowledge:
It models prior information about the fact that the data $x$ is connected to
the parameters $\theta$ in some way. Without this prior knowledge, learning is
impossible: there has to be some relationship between the parameters and the
data, otherwise new data tells you about nothing but itself.

When specific data $x^*$ have been obtained, our state of knowledge about $\theta$
gets updated from the prior distribution to the posterior distribution
via Bayes' rule:
\begin{eqnarray}
p(\theta|x=x^*) &\propto& p(\theta)p(x|\theta)|_{x=x^*} \\
&=& p(\theta)\mathcal{L}(\theta; x)
\end{eqnarray}
The term $p(\theta)p(x|\theta)|_{x=x^*} = \mathcal{L}(\theta; x)$
is the {\it likelihood function}, which is the 
probability of obtaining the actual data set $x^*$ as a function of the
parameters\footnote{In the case that the sampling distribution is a probability
density, the likelihood is the probability density evaluated at the observed
data. This usually causes no problems, although one should be aware of the
Borel-Kolmogorov paradox \citep{jaynes}.}.
As suggested by the above notation, the likelihood function is obtained from the
sampling distribution with the actual data substituted in and is therefore
a function of the parameters only.

To proceed with the model for inferring catalogs from image data, we must
specify a definite hypothesis space and choices for the prior distribution
and the sampling distribution. These choices are presented and discussed in
Section~\ref{sec:model}.

\section{The Specific Model for Stellar Fields}\label{sec:model}
